---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidycjk

<!-- badges: start -->
[![R-CMD-check](https://github.com/PursuitOfDataScience/tidyckj/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PursuitOfDataScience/tidyckj/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Text tooling in R assumes that words are separated by whitespace, and Chinese,
Japanese and Korean writing does not oblige.

tidytext splits on whitespace. Chinese and Japanese don't use whitespace. Give
tidytext an English sentence and you get words; give it 我今天很開心 and you get
either one undifferentiated blob or six isolated characters. Neither is right —
開心 ("happy") is one word of two characters, and splitting it destroys the
signal you were trying to measure.

**tidycjk** tells you what script and language a text column is in, how much of
it is CJK, which characters it contains, how wide it will print, and normalises
the width variants away — as tidy verbs that return tibbles.

## Installation

```r
# install.packages("pak")
pak::pak("PursuitOfDataScience/tidyckj")
```

## The tidy layer

Two verbs take `(data, column)` and return a tibble, so they drop into a dplyr
pipeline.

```{r summary}
library(tidycjk)

posts <- data.frame(
  id = 1:5,
  text = c(
    "我今天很開心",           # Chinese
    "こんにちは、元気ですか",   # Japanese, with kana
    "안녕하세요",              # Korean
    "東京都",                 # Japanese written without kana
    "no CJK here at all"
  )
)

cjk_summary(posts, text)
```

`prop_with_cjk` answers "how many of these rows are CJK at all"; `mean_ratio`
answers "how CJK are they". A corpus of English with one stray ideograph per
row scores high on the first and near zero on the second.

```{r counts}
cjk_char_counts(posts, text)
```

## Which language is this?

```{r language}
cjk_script(posts$text)
cjk_detect_language(posts$text)
```

Row 4 is `NA`, and that is the interesting one. 東京都 is Tokyo Metropolis —
Japanese — but it is written entirely in Han characters, and nothing in the
script distinguishes it from Chinese. Any library that answers `"chinese"`
there is guessing, and it will be wrong on Japanese input in a way you cannot
detect downstream. `tidycjk` returns `NA`.

If your corpus is known to be Chinese, ask for the guess explicitly, so that
the assumption is visible in the script rather than buried in a default:

```{r language-optin}
cjk_detect_language(posts$text, han_only = "chinese")
```

## Display width

`nchar()` counts characters. A terminal counts columns, and a CJK character
takes two of them — which is why every console table containing CJK text comes
out ragged.

```{r width}
labels <- c("中文", "abcd", "日本語")

nchar(labels)      # all look comparable
cjk_width(labels)  # they are not
```

Pad and truncate by width and the columns line up:

```{r layout}
cat(paste0("|", cjk_pad(labels, 8), "|"), sep = "\n")

cjk_truncate("我今天很開心", 8)
```

## Fullwidth and halfwidth

The usual advice is to run `NFKC`. `NFKC` does fix width, and it also rewrites
ligatures, Roman numerals, circled numbers and compatibility ideographs.
`to_halfwidth()` changes width and nothing else.

```{r normalize}
to_halfwidth("１２３")          # fullwidth digits that would not parse
as.numeric(to_halfwidth("１２３"))

to_halfwidth("½ Ⅸ ①")          # NFKC would mangle all three; this leaves them
```

Halfwidth katakana writes a voiced syllable as two code points. Composing them
into one is the difference between text that matches a literal and text that
silently doesn't:

```{r katakana}
x <- "ｶﾞ"          # U+FF76 U+FF9E — two code points
nchar(x)
nchar(to_halfwidth(x))          # one: U+30AC
to_halfwidth(x) == "ガ"
```

## The vector layer

Everything above has a stringr-style counterpart on plain character vectors:
`has_cjk()`, `cjk_script()`, `cjk_detect_language()`, `cjk_ratio()`,
`cjk_width()`, `cjk_pad()`, `cjk_truncate()`, `to_halfwidth()` and
`to_fullwidth()`. All are vectorised, propagate `NA`, and return zero-length
output for zero-length input.

`cjk_blocks()` exports the block table the whole package is built on, so the
definition of "CJK" can be read rather than guessed at.

```{r blocks}
head(cjk_blocks(), 8)
```

## Related work

| Need | Use |
|---|---|
| Display width, width-aware padding | `stringi::stri_width()`, `stringi::stri_pad()` — `cjk_width()` and `cjk_pad()` wrap these |
| Chinese segmentation | `jiebaR` |
| Pinyin | `pinyin`, `hanyupinyin` |
| Traditional/simplified conversion | `tmcn`; `OpenCC` outside R for phrase-level accuracy |
| Japanese utilities | `zipangu`, `Nippon` |

## License

GPL (>= 3).
