---
title: "Romanisation, script conversion and kana"
author: 'Youzhi Yu<br><span style="font-weight: 400; font-size: 0.85em;">University of Chicago</span>'
bibliography: tidycjk.bib
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Romanisation, script conversion and kana}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, message = FALSE, warning = FALSE, comment = "#>"
)
```

```{r setup}
library(tidycjk)
```

## Four conversions, four different levels of trust

This vignette covers the transforms that change *what characters are there*
rather than how wide they are. They are grouped together because they share
an implementation — ICU's transliterators, via **stringi** [@icu;
@gagolewski2022stringi] — and they should not be trusted equally.

| Conversion | Reversible? | Context-free? | Trust |
|---|---|---|---|
| `to_hiragana()` / `to_katakana()` | within one syllabary | yes | complete |
| `cjk_jamo()` / `cjk_compose_jamo()` | returns NFC | yes | complete |
| `cjk_simplify()` | no | context-aware | good |
| `cjk_traditionalize()` | no | context-aware | good on characters, blind to regional vocabulary |
| `cjk_romanize()` | no | no | Chinese yes, Japanese no |

The rest of this vignette is why.

## Kana: safe, with one thing to know

Hiragana and katakana encode the same sounds. The mapping is one to one and
carries no context, so the answer never depends on the surrounding words.

```{r kana}
to_hiragana("カタカナ")
to_katakana("ひらがな")

to_hiragana(to_katakana("ひらがな"))
```

That round trip is exact because the input was already in one syllabary. On
mixed text it is not, and cannot be: converting to a single syllabary erases
the distinction between the two, and that distinction carries meaning —
katakana marks loanwords, onomatopoeia and emphasis.

```{r kana-mixed-roundtrip}
x <- "アか"                      # one katakana, one hiragana
to_hiragana(x)                    # both become hiragana
to_katakana(to_hiragana(x))       # ... and cannot be told apart again
```

So the conversion is safe in the sense that matters — it never guesses — but
run it in one direction, as a normalisation, rather than expecting to undo
it.

Kanji, Latin and punctuation are untouched, so it is safe on a mixed column:

```{r kana-mixed}
to_katakana("日本語のtext です")
```

This is the normalisation you want before grouping Japanese text, where the
same word is often written either way for emphasis.

Halfwidth katakana is handled too, and composed on the way through. `ｶﾞ` is
two code points — the kana and a combining voiced mark — and comes back as
the single character `ガ`:

```{r kana-width}
to_hiragana("ｶﾀｶﾅ")
to_katakana("ｶﾞ")
nchar(to_katakana("ｶﾞ"))
```

## Hangul jamo: safe

A modern Hangul syllable is built from a leading consonant, a vowel and an
optional trailing consonant, and Unicode encodes all 11,172 combinations
precomposed. The relationship is arithmetic —
`SIndex = (LIndex * 21 + VIndex) * 28 + TIndex` — so the decomposition is
exact rather than tabulated: no table can be out of date and no syllable is
missed.

```{r jamo}
cjk_jamo("한글")
```

```{r jamo-roundtrip}
back <- vapply(cjk_jamo("한글"), paste, character(1), collapse = "")
cjk_compose_jamo(back)
```

One caveat on the round trip: `cjk_compose_jamo()` is normalisation form C,
so it composes everything composable, not only the jamo it was given. Text
that was already in NFC — almost anything you will meet — comes back
unchanged. Text that was not comes back normalised:

```{r jamo-nfc}
x <- "e\u0301"                        # e followed by a combining acute
cjk_compose_jamo(paste(cjk_jamo(x)[[1]], collapse = ""))
```

That is `é` as a single code point. The guarantee is that the round trip
returns `stringi::stri_trans_nfc(x)`, which is `x` whenever `x` is already
NFC.

Jamo are the right unit for questions the syllable hides — which initial
consonants a corpus favours, or whether two spellings differ only in a final
consonant. `한` is one character to `cjk_width()` and three jamo here:

```{r jamo-count}
cjk_width("한")
lengths(cjk_jamo("한"))
```

## Simplified and traditional Han: right on characters, blind to vocabulary

```{r han}
cjk_simplify("漢字")
cjk_traditionalize("汉字")
```

It is better than it looks. Simplified → traditional is genuinely
one-to-many — simplified `发` is `發` ("to send") or `髮` ("hair"), and `干`
is `乾`, `幹` or `干` — and ICU resolves these **from context**, not
character by character:

```{r han-context}
cjk_traditionalize(c("头发", "发送"))   # hair, send
cjk_traditionalize(c("干净", "树干"))   # clean, tree trunk
cjk_traditionalize(c("后天", "皇后"))   # after, empress — 后 splits both ways
```

All six are right. Over fourteen such pairs, including `面条`/`面对` and
`里面`/`公里`, every one came out correct.

What ICU does *not* do is substitute regional vocabulary. The two standards
differ in the words they use, not only in glyph shape, and that is lexical:

```{r han-limit}
cjk_traditionalize(c("软件", "鼠标", "网络"))
```

Those are the correct characters and the wrong words in Taiwan, where the
terms are `軟體`, `滑鼠` and `網路`. No character mapping can reach that, and
it is what [OpenCC](https://github.com/BYVoid/OpenCC) [@opencc] and its
regional configurations exist for. Use these functions for script
conversion; reach for OpenCC when you need Taiwanese or Hong Kong idiom.

## Romanisation: depends entirely on the language

```{r romanize}
cjk_romanize("中文")
cjk_romanize("中文", ascii = TRUE)
```

For Chinese this is pinyin, with tone marks unless you ask for ASCII, and it
is good. `ascii = TRUE` gives a sortable, greppable key for a CJK column,
which is often the real reason to romanise at all.

```{r romanize-scripts}
cjk_romanize(c("こんにちは", "안녕하세요"))
```

Three caveats, in increasing order of severity.

**Han readings are per character.** A character with several readings gets
ICU's preferred one regardless of the word it appears in.

**No boundary between scripts.** ICU spaces Han syllables but does not insert
a break where the script changes:

```{r romanize-mixed}
cjk_romanize("漢字カナ")
```

Segment first with `cjk_segment()` if you need word boundaries in the output.

**Japanese kanji come back as Chinese.** This is the severe one. ICU routes
every Han character through pinyin regardless of the surrounding language, so
`日本語` romanises to `rì běn yǔ` rather than `nihongo`, and `私` to `sī`
rather than `watashi`:

```{r romanize-ja}
cjk_romanize("私は日本語を話します")
```

That is not an approximation of the Japanese reading — it is the Chinese one.
The kana in the same string romanise correctly, which makes the output look
plausible at a glance and is exactly what makes it dangerous. Note too that
the particle `は` gives `ha`, which is how it is written rather than how it
is said.

For Japanese, use a morphological analyser that carries readings — **gibasa**
binds MeCab and returns them [@kudo2004mecab; @gibasa]. `cjk_romanize()` is
for Chinese, for kana, and for producing an ASCII sort key.

## Width forms are a separate concern

`to_halfwidth()` and `to_fullwidth()` change width and nothing else. The usual
advice for fullwidth text is an `NFKC` pass, which does fix width — and also
rewrites ligatures, Roman numerals, circled numbers and compatibility
ideographs, none of which you asked for:

```{r width-forms}
as.numeric(to_halfwidth("１２３"))

to_halfwidth("½ Ⅸ ①")   # NFKC rewrites all three; this leaves them
```

The one composition it does perform is voiced halfwidth katakana, where a
single syllable is written as two code points. Composing them is the
difference between text that matches a literal and text that silently does
not:

```{r voiced}
x <- "ｶﾞ"
nchar(x)
nchar(to_halfwidth(x))
to_halfwidth(x) == "ガ"
```

## References
