---
title: "Width, padding, truncation and wrapping"
author: 'Youzhi Yu<br><span style="font-weight: 400; font-size: 0.85em;">University of Chicago</span>'
bibliography: tidycjk.bib
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Width, padding, truncation and wrapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, message = FALSE, warning = FALSE, comment = "#>"
)
```

```{r setup}
library(tidycjk)
```

## One character is not one column

This is the whole vignette in one comparison:

```{r basic}
labels <- c("中文", "abcd", "日本語")

nchar(labels)      # characters
cjk_width(labels)  # terminal columns
```

`中文` is two characters and four columns. Unicode Annex #11 assigns every
character an East Asian Width, and the values `Wide` and `Fullwidth` mean two
columns in a monospaced terminal [@uax11]. `cjk_width()` reports that number.

Everything that goes wrong with CJK in a console follows from code that used
the first number where it needed the second.

```{r ragged}
pad_by_char <- function(x, n) paste0(x, strrep(" ", n - nchar(x)))

cat(paste0("|", pad_by_char(labels, 8), "|"), sep = "\n")
```

Three lines that should be the same width, and are not. Now with the column
count:

```{r aligned}
cat(paste0("|", cjk_pad(labels, 8), "|"), sep = "\n")
```

## `cjk_pad()`

Pads to a width in columns. `side` names the side the padding goes on, and
`pad` must itself be one column wide — a two-column pad character cannot hit
an odd target, so it is rejected rather than allowed to overshoot.

```{r pad}
cjk_pad("中文", 10)
cjk_pad("中文", 10, side = "left")
cjk_pad("中文", 10, side = "both")
cjk_pad("中文", 10, pad = ".")
```

A width already exceeded is left alone rather than truncated — padding and
truncation are separate verbs on purpose:

```{r pad-over}
cjk_width(cjk_pad("日本語テキスト", 4))
```

## `cjk_truncate()`

Cuts to a width, again in columns, appending an ellipsis that is itself
counted against the budget:

```{r truncate}
cjk_truncate("我今天很開心", 8)
cjk_width(cjk_truncate("我今天很開心", 8))
```

The cut never lands inside a character. Because a CJK character is two
columns, an odd budget can leave one column unused, and leaving it empty is
the only correct option:

```{r truncate-odd}
cjk_truncate("中文中文", 5, ellipsis = "")
cjk_width(cjk_truncate("中文中文", 5, ellipsis = ""))
```

## `cjk_wrap()`

Breaks a string into lines that each fit the budget. This is where counting
columns stops being sufficient, because *where* you are allowed to break is
its own question.

```{r wrap}
x <- "我今天很開心，因為天氣非常好而且朋友來看我"
cat(cjk_wrap(x, 14))
```

Every line is inside the budget:

```{r wrap-check}
cjk_width(strsplit(cjk_wrap(x, 14), "\n", fixed = TRUE)[[1]])
```

### Why this is not "split every N columns"

CJK text is *mostly* breakable between any two characters — there are no
spaces to break at — but not everywhere. Unicode Annex #14 defines the
positions [@uax14], and several are forbidden:

* a closing bracket `）` may not begin a line;
* a full stop `。` may not begin a line;
* an opening bracket `（` may not end one.

A naive chunker violates all three. `cjk_wrap()` uses ICU's implementation of
the algorithm, so the break positions are the ones a typesetter would choose:

```{r wrap-punct}
cat(cjk_wrap("他說（今天天氣很好）。我們去公園散步。", 12))
```

Notice that no line begins with `。` or `）`.

### One rule that is not universal

Those three hold whatever the locale. A fourth — whether a small kana may be
separated from the character it follows — does **not**. Annex #14 defines
strict, normal and loose line-breaking styles and which applies is tailored
per locale: under `"en"` a small kana never begins a line, and under `"ja"`,
which uses the looser style, it may. Name the style when the output has to be
reproducible:

```{r wrap-style}
kana <- "きょうはとてもいいてんきですっしゃちょ"
cat(cjk_wrap(kana, 6, locale = "ja@lb=strict"))
```

Latin runs inside the same string still break on spaces, so mixed text
behaves the way you would expect from both sides:

```{r wrap-mixed}
cat(cjk_wrap("tidycjk 讓中日韓文字在終端機裡對齊 and it handles Latin too", 20))
```

### Indents

```{r wrap-indent}
cat(cjk_wrap(x, 14, exdent = 2))
```

## The shared contract

All three layout verbs behave identically at the edges, which is what makes
them safe inside a pipeline:

```{r contract}
cjk_pad(c("中", NA), 6)
cjk_truncate(character(0), 6)
cjk_wrap(NA, 6)
```

A `width` that is `Inf` or outside integer range is an error rather than a
silent `NA`. `cjk_pad(x, Inf)` is a plausible way to write "do not truncate",
and it used to return missing values with only a coercion warning:

```{r contract-error, error = TRUE}
cjk_pad("中文", Inf)
```

Ragged recycling is an error too, naming both lengths, because a plausible
vector of the wrong length is worse for a layout function than a refusal:

```{r contract-recycle, error = TRUE}
cjk_pad(c("a", "b", "c"), c(2, 3))
```

## Ambiguous width

Some characters — Greek, Cyrillic, a few punctuation marks — are `Ambiguous`
in Annex #11: two columns in a legacy East Asian font, one otherwise. There is
no context-free answer, and ICU has changed its treatment more than once.
`cjk_width()` reports what ICU reports on your machine. If your output must
align in a specific terminal, measure there rather than assuming:

```{r ambiguous}
cjk_width(c("α", "→", "±"))
```

## References
